Local Spectral Clustering of Density Upper Level Sets

### 2019-01-12 icml/icml_local_density_clustering.tex
finished results, abstract, and discussion; added outline for examples

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\text{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}

@@

%%% Sets
\newcommand{\Cset}{\mathcal{C}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Asig}{\Aset_{\sigma}}

@@

@@

\theoremstyle{aldenthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

@@

\begin{abstract}
\end{abstract}

@@

Roughly speaking, spectral clustering techniques first embed the data $\Xbf$ using the spectrum of the graph Laplacian matrix and subsequently 
use this \emph{spectral embedding} to find a clustering of the data.
When applied to large graphs (or large point clouds) classical global spectral 
methods can be computationally cumbersome and 
can be insensitive to the local geometry of the distribution of the samples
\citep{mahoney2012,leskovec2010}.
This in turn has led to the investigation of local spectral algorithms \citep{spielman2013,anderson2006,leskovec2010}
which leverage locally-biased spectra computed using random walks around 
a user-specified seed node.

@@

For $\Wbf$ as before, let the graph $G  = (V,E)$, with vertices $V = \set{v_1, \ldots, v_n}$ corresponding to the $n$ rows of $\Abf$, and (possibly weighted) edges $E = \set{(v_i,v_j,\Abf_{ij}): 1 \leq i < j \leq n, \Abf_{ij} > 0}$ (Since $\Abf$ is symmetric, $G$ is an undirected graph; also, by convention, we preclude self-loops, hence $i \neq j$). There are many \cite{yang2015,fortunato2010} graph-theoretic measures which assess the cluster quality of a subset $S \subseteq V$ (or more generally the quality of a partition $S_1 \cup \ldots \cup S_m = V$, for $m \geq 2$.)

@@

We will assess the external connectivity of a subset $S \subseteq V$ through its normalized cut. The cut of $S$ is
\begin{equation*}
\cut(S; G) := \sum_{u \in S} \sum_{v \in S^c} \1\left((u,v) \in E \right)
\end{equation*}
-- where $S^c = V / S$ is the complement of $S$ in $V$ -- and the volume of $S$ is
\begin{equation*}
\vol(S; G) := \sum_{u \in S} \sum_{v \in V} \1\left((u,v) \in E \right).
\end{equation*}
Then, the \textit{normalized cut} of $S$ is given by
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) := \frac{\cut(S; G)}{ \min\left\{\vol(S; G), \vol(S^c; G) \right\} }
\end{equation} 
Intuitively, a set with low \textit{normalized cut} has many more edges which do not cross the cut than edges which do cross the cut.

Given $S \subseteq V$, the subgraph induced by $S$ is given by $G[S] = (S, E_S)$, where $(u,v) \in E_S$ if both $u$ and $v$ are in $S$ and $(u,v) \in E_S$. Letting $\abs{S} = m$, $\Abf_S$ then denotes the $m \times m$ adjacency matrix representation of $G[S]$; similarly, $\Dbf_S$ is the diagonal degree matrix with entries $(\Dbf_{S})_{ii} = \sum_{j: v_j \in S} \Abf_{ij}$, and $\Wbf = \Dbf_S^{-1} \Abf_S$ is the corresponding random walk matrix (again, only over $G[S]$.)

Our internal connectivity parameter $\Psi(S)$ will capture the time it takes for the random walk governed by $\Wbf_S$ to mix (that is, approach a stationary distribution) uniformly over $S$. Denoting the stationary distribution $\widetilde{\pibf} = (\widetilde{\pi}_{1}, \ldots, \widetilde{\pi}_{m})$, the \textit{relative pointwise mixing time} $\tau_{\infty}(S; G[S])$ of the random walk over $G[S]$ defined by $\Wbf_S$ is the smallest integer $t_0 > 0$ such that for all $v = v_i,v' = v_j \in S$:
\begin{equation*}
\abs{\frac{e_v \Wbf_S^t - \widetilde{\pi}_{j}}{\widetilde{\pi}_{j}}} \leq \frac{1}{4}
\end{equation*}
for all $t > t_0$
\footnote{Given a seed node $v$ and and a random walk defined by transition probability matrix $\mathbf{P}$, the rotation $e_v \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$ steps.} (Of course, given the definition of $\Wbf_S$ it is well known that the stationary distribution $\pibf_S$ will be defined by $\widetilde{\pi}_{i} = (\Dbf_{S})_{ii} / \vol(S; G[S])$.

@@

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no less than $\Psi$, we will refer to it as a $(\Phi,\Psi)$-cluster. Both local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been shown to output clusters (or partitions) which provably satisfy approximations to the optimal $(\Phi, \Psi)$-cluster (or partition), where the optimization is carried out over the graph $G$. \footnote{In the case of \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the conductance, i.e. the minimum normalized cut within the subgraph $G[S]$. See Theorem 3.1 for details; however, note that $\phi^2 / \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a lower bound on $\Psi$.}

\paragraph{Personalized PageRank.}
As mentioned previously, global algorithms which find spectral cuts may be computationally infeasible for large graphs; in this setting, local algorithms may be preferred or even required. We will restrict our attention in particular to one such popular algorithm: \textit{personalized PageRank} (PPR). The personalized PageRank algorithm was first introduced by \cite{haveliwala2003} and variants of this algorithm have been studied further in several recent works \citep{spielman2011,spielman2014,zhu2013,anderson2006,mahoney2012}.

The random walk matrix $\Wbf$ over the graph $G = (V,E)$ with associated adjacency matrix $\Abf$ is defined as in (\ref{eqn: random_walk_laplacian}). PPR is then defined with respect to the following inputs: a user-specified seed node $v_i \in V$ , and  $\alpha \in [0,1]$ a teleportation parameter. Letting $v = v_i$ for notational simplicity, and $e_{v}$ be the indicator vector for $v$ (meaning $e_v$ has a $1$ in the $i$th location and $0$ everywhere else), the \textit{PPR vector} is given by the recursive formulation
\begin{equation}
\label{eqn: ppr_vector}
\pbf(v,\alpha;G) := \alpha \ebf{v} + (1 - \alpha) \pbf(v,\alpha;G) \Wbf
\end{equation}
We note in passing that, for $\alpha > 0$, the vector $\pbf(v,\alpha;G)$ can be well-approximated by a simple local computation (of a random walk with restarts at the node $v$.) We also point out that, from a density clustering standpoint, since density clusters are inherently local, using the PPR algorithm eases the analysis, and as we will observe in the sequel our analysis requires fewer global regularity conditions relative to more classical global spectral algorithms.

To compute a cluster $\widehat{C} \subset V$ using the PPR vector, we will take sweep cuts of $\pbf(v, \alpha; G)$. Denote the entries of $\pbf(v, \alpha; G)$ by $\pbf(v, \alpha; G) = (p_1, \ldots, p_n)$, and let the \textit{stationary distribution} $\pibf$ of the random walk defined by $\Wbf$ be given by
\begin{equation*}
\pibf = (\pi_1, \ldots, \pi_n), ~~~ \pi_j := \frac{\Dbf_{jj}}{\vol(V; G)}.
\end{equation*}
Then, for a number $\beta \geq 0$, the sweep cut $S_\beta$ is
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u_j \in V: p_j > \beta \pi_j}.
\end{equation}
One such sweep cut will be our chosen cluster $\widehat{C}$.  
(We delay formal introduction of the local clustering algorithm we analyze until \textcolor{red}{Section 2}, after we have given a method for forming a graph over the data $\Xbf$.)

@@

It is worth pointing out that in this context, continuous analogues to (for instance) normalized cut have been defined, over the data-manifold rather than the graph, and convergence of finite sample graph-theoretic functionals to their continuous counterparts has been shown
\cite{garciatrillos16, arias-castro12, maier11}.
However these continuous analogues are not always easily interpretable -- and their corresponding minimizers not always easily identifiable -- for the particular density function under consideration. Of course, relating these partitions to the arguably more simply defined high density clusters can be also challenging in general. \textcolor{red}{Intuitively, however, under the right conditions such high-density clusters should have more edges within themselves than to the remainder of the graph.} We formalize this intuition next.

@@

Hereafter, we consider the \textit{uniform kernel function} for a fixed $r > 0$,
\begin{equation}
k(x,x') = \1(\norm{x - x'} \leq r)
\end{equation}
and the associated \textit{neighborhood graph} 
\begin{equation}
\label{eqn: neighborhood_graph}
G_{n,r} = (\Xbf, E_{n,r}), \text{  $(x_i,x_j) \in E_{n,r}$ if $k(x_i,x_j) = 1$}
\end{equation}
For a given high density cluster $\Cset \subseteq \Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \textit{empirical density cluster}. We now introduce a notion of consistency for the task of density cluster estimation:

\begin{definition}[Consistent density cluster estimation]
	For an estimator $\widehat{\Cset}_n \subset \Xbf$, and any $\Cset, \Cset' \in \Cbb_f(\lambda),$ we say $\widehat{\Cset}_n$ is a consistent estimator of $\Cset$ if the following statement holds: as the sample size $n \to \infty$, each of the following
	\begin{equation}
	\label{eqn: consistent_density_cluster_recovery}
	\Cset[\Xbf] \subseteq \widehat{\Cset}_n, ~\mathrm{ and }~ \widehat{\Cset}_n \cap \Cset'[\Xbf] = \emptyset
	\end{equation}
	occur with probability tending to $1$.

@@

	\item 
	We show these bounds on the graph connectivity criteria have algorithmic consequences personalized PageRank. An immediate consequence of Theorems \ref{thm: conductance_upper_bound} and \textcolor{red}{Theorem 2}, along with the previous work of \cite{zhu2013}, is to yield an upper bound on the normalized cut of the set $\widehat{\Cset}_n$ output by Algorithm \ref{alg: ppr}, as well as upper bounding the symmetric set difference between $\widehat{\Cset}_n$ and $\Cset[\Xbf]$. Furthermore, in \textcolor{red}{Section 4} we show that a careful analysis of the form typical to local clustering algorithms yields \textcolor{red}{Theorem 4}, which states that Algorithm \ref{alg: ppr}, properly initialized, performs consistent density cluster estimation in the sense of (\ref{eqn: consistent_density_cluster_recovery}).
\end{enumerate}

\paragraph{Organization.}
In Section \ref{section: everything_else}, we provide some example density functions, to clarify the relevance of our results. In \textcolor{red}{Section 6}, we show empirical performance of the PPR algorithm, which demonstrates that violations of the geometric conditions we set out in Section \ref{sec: measures_of_cluster_quality} manifestly impact density cluster recovery (i.e. the conditions are not superfluous), before concluding in \textcolor{red}{Section 7}. First, however, we summarize some related work.

@@

Let $B(x,r) = \set{y \in \Rd: \norm{y - x} \leq r}$ be a closed ball of radius $r$ around the point $x$.  Given a set $\Aset \subset \Rd$, and a number $\sigma > 0$, define the set $\Asig = \Aset + B(0,\sigma) = \set{y \in \Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma}$. The assumptions we require are as follows: 
\begin{enumerate}[label=(A\arabic*)]
	\item 
	\label{asmp: min_density}
	\textit{Minimum density:} For numbers $\sigma, \lambda_{\sigma} > 0$, a set $\Aset \subset \Rd$ is said to be \textit{$(\sigma, \lambda_{\sigma})$-regular} if
	\begin{equation}
	\inf_{x \in \Asig} f(x) = \lambda_{\sigma}
	\end{equation} 
	%and 
	%\begin{equation*}
	%\frac{\diam \Asig}{\sigma} \leq \mu
	%\end{equation*}
	%where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

	\item 
	\label{asmp: low_noise_density}
	\textit{Low noise density:} For numbers $\gamma, \sigma > 0$, and a set $\Aset \subset \Rd$, the density function $f$ is said to be \textit{$(\gamma, \sigma)$- low noise around $\Aset$} if there exists a constant $c_1 > 0$ such that for all $x \in \Rd$ with $0 < \rho(x, \Asig) \leq \sigma$,
	\begin{equation*}
	\inf_{x' \in \Asig} f(x') - f(x) \geq c_1 \rho(x, \Asig)^{\gamma},
	\end{equation*}
	where $\rho(x,\Asig) = \min_{x_0 \in \Asig} \norm{x - x_0}$.

	\item
	\label{asmp: cluster_separation}
	\textit{Cluster separation:}
	For $\lambda,\sigma > 0$, $\Cset \in \Cbb_f(\lambda)$ is said to be \textit{$\sigma$-well separated} if for all $\Cset' \in \Cbb_f(\lambda)$,
	\begin{equation*}
	\rho(\Cset,\Cset') > \sigma.
	\end{equation*}
	where $\rho(\Cset,\Cset') = \min_{x \in \Cset} \rho(x,\Cset')$,
\end{enumerate}

Assumptions \ref{asmp: min_density}-\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf]; G_{n,r})$, whereas \ref{asmp: min_density} and \textcolor{red}{(A4)} are necessary to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that \ref{asmp: min_density} and \ref{asmp: cluster_separation} combined are similar to the $(\sigma,\epsilon)$-saliency of \cite{chaudhuri2010}, a standard density clustering assumption, while \ref{asmp: low_noise_density} is seen in, for instance, \cite{singh2009}, (as well as many other works on density clustering and level set estimation.) Further, it is worth highlighting that these assumptions are all local in nature, a benefit of studying a local algorithm such as PPR.

\begin{definition}
	For $\lambda > 0$ we say $\Cset \in \mathbb{C}_f(\lambda)$ is a $(\lambda_{\sigma}, \gamma, \sigma)$-well conditioned cluster if it satisfies Assumptions \ref{asmp: min_density}-\ref{asmp: cluster_separation}, for some $\lambda_{\sigma}, \gamma, \sigma > 0$.
\end{definition}
We note that $\sigma$ plays dual roles here, both in effect precluding arbitrarily narrow and long clusters in \ref{asmp: min_density} and arbitrarily flat densities in \ref{asmp: low_noise_density}.

@@

Algorithm \ref{alg: ppr} will be the simple procedure we analyze. It will take as input the data $\Xbf$ along with user-specified parameters $r, \alpha$, $\vol_0$, and $v \in \Xbf$, and perform the following steps:

>>>>

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\mathbf{\Phi}}
\newcommand{\Psibf}{\mathbf{\Psi}}

@@

%%% Sets
\newcommand{\Cset}{\mathcal{C}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Asig}{\Aset_{\sigma}}
\newcommand{\Csig}{\Cset_{\sigma}}

@@

%%% Algorithm notation
\newcommand{\ppr}{{\sc PPR}}
\newcommand{\pprspace}{{\sc PPR~}}

@@

\theoremstyle{aldenthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

@@

\begin{abstract}
	Spectral clustering methods are 
	a family of popular non-parametric clustering tools.
	Recent works have proposed and analyzed 
	\emph{local} spectral methods, which 
	extract clusters using
	locally-biased random walks around a user-specified seed node, and are known to have worst-case guarantees for certain graph-based measures of cluster quality.
	In contrast to existing works, 
	we analyze the personalized PageRank (\ppr) algorithm in the classical statistical learning setup,
	where we obtain samples from an unknown distribution, and aim to identify connected
	regions of high-density (density clusters).
	With respect to a neighborhood graph formed over these samples, we provide guarantees for a pair of cluster quality criteria evaluated on empirical analogues to these density clusters. As a result, the \pprspace algorithm is shown to 
	extract sufficiently salient density clusters, and we 
	analyze the sample-complexity of the algorithm.
\end{abstract}

@@

Roughly speaking, spectral clustering techniques first embed the data $\Xbf$ using the spectrum of the graph Laplacian, and subsequently 
use this \emph{spectral embedding} to find a clustering of the data.
When applied to large graphs (or large point clouds) classical global spectral 
methods can be computationally cumbersome and 
insensitive to the local geometry of the distribution of the samples
\citep{mahoney2012,leskovec2010}.
This in turn has led to the investigation of local spectral algorithms \citep{spielman2013,anderson2006,leskovec2010}
which leverage locally-biased spectra computed using random walks around 
a user-specified seed node.

@@

For $\Abf$ as before, consider the graph $G  = (V,E)$, with vertices $V = \set{v_1, \ldots, v_n}$ corresponding to the $n$ rows of $\Abf$, and (possibly weighted) edges $E = \set{(v_i,v_j,\Abf_{ij}): 1 \leq i < j \leq n, \Abf_{ij} > 0}$ (Since $\Abf$ is symmetric, $G$ is an undirected graph; also, by convention, we preclude self-loops, hence $i \neq j$). There are many \cite{yang2015,fortunato2010} graph-based measures which assess the cluster quality of a subset $S \subseteq V$ (or more generally the quality of a partition $S_1 \cup \ldots \cup S_m = V$, for $m \geq 2$.)

@@

We will assess the external connectivity of a subset $S \subseteq V$ through its normalized cut. The cut of $S$ is
\begin{equation*}
\cut(S; G) := \sum_{u \in S} \sum_{v \in S^c} \1\left((u,v) \in E \right)
\end{equation*}
-- where $S^c = V / S$ is the complement of $S$ in $V$ -- and the volume of $S$ is
\begin{equation*}
\vol(S; G) := \sum_{u \in S} \sum_{v \in V} \1\left((u,v) \in E \right).
\end{equation*}
Then, the \emph{normalized cut} of $S$ is given by
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) := \frac{\cut(S; G)}{ \min\left\{\vol(S; G), \vol(S^c; G) \right\} }
\end{equation} 
Intuitively, a set with low \textit{normalized cut} has many more edges which do not cross the cut than edges which do cross the cut.

Given $S \subseteq V$, the subgraph induced by $S$ is given by $G[S] = (S, E_S)$, where $(u,v) \in E_S$ if both $u$ and $v$ are in $S$ and $(u,v) \in E_S$. Letting $\abs{S} = m$, $\Abf_S$ then denotes the $m \times m$ adjacency matrix representation of $G[S]$; similarly, $\Dbf_S$ is the diagonal degree matrix with entries $(\Dbf_{S})_{ii} = \sum_{j: v_j \in S} \Abf_{ij}$, and $\Wbf = \Dbf_S^{-1} \Abf_S$ is the corresponding random walk matrix (again, defined only over $G[S]$.)

Our internal connectivity parameter $\Psi(S)$ will capture the time it takes for the random walk governed by $\Wbf_S$ to mix (that is, approach a stationary distribution) uniformly over $S$. Denoting the stationary distribution of this random walk by $\widetilde{\pibf} = (\widetilde{\pi}_{1}, \ldots, \widetilde{\pi}_{m})$, the \emph{relative pointwise mixing time} $\tau_{\infty}(S; G[S])$ is defined to be: the smallest integer $t_0 > 0$ such that for all $v = v_i,v' = v_j \in S$:
\begin{equation*}
\abs{\frac{e_v \Wbf_S^t - \widetilde{\pi}_{j}}{\widetilde{\pi}_{j}}} \leq \frac{1}{4}
\end{equation*}
for all $t > t_0$
\footnote{Given a starting node $v$ and and a random walk defined by transition probability matrix $\mathbf{P}$, the rotation $e_v \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$ steps.} (Of course, given the definition of $\Wbf_S$ it is well known that the stationary distribution $\pibf_S$ will be defined by $\widetilde{\pi}_{i} = (\Dbf_{S})_{ii} / \vol(S; G[S])$.

@@

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no less than $\Psi$, we will refer to it as a $(\Phi,\Psi)$-cluster. Both local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been shown to output clusters (or partitions) which provably satisfy approximations to the optimal $(\Phi, \Psi)$-cluster (or partition) for a given graph $G$. \footnote{In the case of \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the conductance, i.e. the minimum normalized cut within the subgraph $G[S]$. See Theorem 3.1 for details; however, note that $\phi^2 / \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a lower bound on $\Psi$.}

\paragraph{Personalized PageRank.}
As mentioned previously, global algorithms which find spectral cuts may be computationally infeasible for large graphs; in this setting, local algorithms may be preferred or even required. We will restrict our attention in particular to one such popular algorithm: \emph{personalized PageRank} (\ppr). The personalized PageRank algorithm was first introduced by \cite{haveliwala2003} and variants of this algorithm have been studied further in several recent works \citep{spielman2011,spielman2014,zhu2013,anderson2006,mahoney2012}.

The random walk matrix $\Wbf$ over the graph $G = (V,E)$ with associated adjacency matrix $\Abf$ is given by (\ref{eqn: random_walk_laplacian}). \pprspace is then defined with respect to the following inputs: a user-specified seed node $v_i \in V$ , and  $\alpha \in [0,1]$ a teleportation parameter. Letting $v = v_i$ for notational simplicity, and $e_{v}$ be the indicator vector for $v$ (meaning $e_v$ has a $1$ in the $i$th location and $0$ everywhere else), the \emph{PPR vector} is given by the recursive formulation
\begin{equation}
\label{eqn: ppr_vector}
\pbf(v,\alpha;G) := \alpha \ebf{v} + (1 - \alpha) \pbf(v,\alpha;G) \Wbf
\end{equation}
We note in passing that, for $\alpha > 0$, the vector $\pbf(v,\alpha;G)$ can be well-approximated by a simple local computation (of a random walk with restarts at the node $v$.) We also point out that, from a density clustering standpoint, since density clusters are inherently local, using the PPR algorithm eases the analysis, and as we will observe in the sequel our analysis requires fewer global regularity conditions relative to more classical global spectral algorithms.

To compute a cluster $\widehat{C} \subset V$ using the \pprspace vector, we will take sweep cuts of $\pbf(v, \alpha; G)$. Denote the entries of $\pbf(v, \alpha; G)$ by $\pbf(v, \alpha; G) = (p_1, \ldots, p_n)$, and let the \emph{stationary distribution} $\pibf$ of the random walk defined by $\Wbf$ be given by
\begin{equation*}
\pibf = (\pi_1, \ldots, \pi_n), ~~~ \pi_j := \frac{\Dbf_{jj}}{\vol(V; G)}.
\end{equation*}
Then, for a number $\beta \geq 0$, the sweep cut $S_\beta$ is
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u_j \in V: p_j > \beta \pi_j}.
\end{equation}
We will choose one such sweep cut to be our cluster estimate $\widehat{C}$.  
(We delay formal introduction of the local clustering algorithm we analyze until \ref{sec: background}, after we have given a method for forming a graph over the data $\Xbf$.)

@@

In this context, continuous analogues to (for instance) normalized cut have been defined, over the data-manifold rather than the graph, and convergence of finite sample graph-theoretic functionals to their continuous counterparts has been shown
\cite{garciatrillos16, arias-castro12, maier11}.
However these continuous analogues are not always easily interpretable -- and their corresponding minimizers not always easily identifiable -- for the particular density function under consideration. Of course, relating these partitions to the arguably more simply defined high density clusters can be also challenging in general. Intuitively, however, under the right conditions such high-density clusters should have more edges within themselves than to the remainder of the graph. We formalize this intuition next.

@@

Hereafter, we consider the uniform kernel function for a fixed $r > 0$,
\begin{equation}
k(x,x') = \1(\norm{x - x'} \leq r)
\end{equation}
and the associated neighborhood graph
\begin{equation}
\label{eqn: neighborhood_graph}
G_{n,r} = (\Xbf, E_{n,r}), \text{  $(x_i,x_j) \in E_{n,r}$ if $k(x_i,x_j) = 1$}
\end{equation}
For a given high density cluster $\Cset \subseteq \Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. We now introduce a notion of consistency for the task of density cluster estimation:

\begin{definition}[Consistent density cluster estimation]
	\label{def: consistent_density_cluster_estimation}
	For an estimator $\widehat{\Cset}_n \subset \Xbf$, and any $\Cset, \Cset' \in \Cbb_f(\lambda),$ we say $\widehat{\Cset}_n$ is a consistent estimator of $\Cset$ if the following statement holds: as the sample size $n \to \infty$, each of the following
	\begin{equation}
	\label{eqn: consistent_density_cluster_recovery}
	\Cset[\Xbf] \subseteq \widehat{\Cset}_n, ~\mathrm{ and }~ \widehat{\Cset}_n \cap \Cset'[\Xbf] = \emptyset
	\end{equation}
	occur with probability tending to $1$.

@@

	\item 
	We show these bounds on the graph connectivity criteria have algorithmic consequences for personalized PageRank. An immediate consequence of Theorems \ref{thm: conductance_upper_bound} and \ref{thm: inverse_mixing_time_lower_bound}, along with the previous work of \cite{zhu2013}, is to yield an upper bound on the normalized cut of the set $\widehat{C}$ output by Algorithm \ref{alg: ppr}, as well as upper bounding the symmetric set difference between $\widehat{C}$ and $\Cset[\Xbf]$. Furthermore, in \ref{sec: consistent_cluster_estimation_with_ppr} we show that a careful analysis of the form typical to local clustering algorithms yields Theorem \ref{thm: consistent_recovery_of_density_clusters}, which states that Algorithm \ref{alg: ppr}, properly initialized, performs consistent density cluster estimation in the sense of (\ref{eqn: consistent_density_cluster_recovery}).
\end{enumerate}

\paragraph{Organization.}
In Section \ref{section: examples}, we provide some example density functions, to clarify the relevance of our results, and empirically demonstrate that violations of the geometric conditions we set out in Section \ref{sec: background} manifestly impact density cluster recovery (i.e. the conditions are not superfluous), before concluding in \ref{sec: discussion}. First, however, we summarize some related work.

@@

Let $B(x,r) = \set{y \in \Rd: \norm{y - x} \leq r}$ be a closed ball of radius $r$ around the point $x$.  Given a set $\Aset \subset \Rd$, and a number $\sigma > 0$, define the $\sigma$-expansion of $\Aset$ to be $\Asig = \Aset + B(0,\sigma) = \set{y \in \Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma}$. We are now ready to give the assumptions, which we state with respect to a density cluster $\Cset \in \Cbb_f(\lambda)$ for some $\lambda > 0$, and expansion parameter $\sigma > 0$:
\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: bounded_density}
	\textit{Bounded density within cluster:} There exist numbers $0 < \lambda_{\sigma} <  \Lambda_{\sigma} < \infty$ such that:
	\begin{equation}
	\label{eqn: bounded_density}
	\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq \Lambda_{\sigma} 
	\end{equation} 
	%and 
	%\begin{equation*}
	%\frac{\diam \Asig}{\sigma} \leq \mu
	%\end{equation*}
	%where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

	\item 
	\label{asmp: low_noise_density}
	\textit{Low noise density:} For some $\gamma > 0 $, there exists a constant $c_1 > 0$ such that for all $x \in \Rd$ with $0 < \rho(x, \Csig) \leq \sigma$,
	\begin{equation*}
	\inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \rho(x, \Csig)^{\gamma},
	\end{equation*}
	where $\rho(x,\Aset) = \min_{x_0 \in \Aset} \norm{x - x_0}$ for $\Aset \subset \Rd$.

	\item
	\label{asmp: cluster_separation}
	\textit{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$,
	\begin{equation*}
	\rho(\Csig,\Csig') > \sigma,
	\end{equation*}
	where $\rho(\Aset,\Aset') = \min_{x \in \Aset} \rho(x,\Aset')$ for $\Aset' \subset \Rd$.

	\item
	\label{asmp: cluster_diameter}
	\textit{Cluster diameter:}
	There exists $D < \infty$ such that for all $x, x' \in \Csig$:
	\begin{equation*}
	\norm{x - x'} \leq D.
	\end{equation*}
\end{enumerate}

We note that $\sigma$ plays several roles here, precluding arbitrarily narrow clusters and long clusters in \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter}, flat densities around the level set in \ref{asmp: low_noise_density}, and poorly separated clusters in \ref{asmp: cluster_separation}.

Assumptions \ref{asmp: bounded_density}-\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf]; G_{n,r})$, whereas \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} are necessary to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the lower bound on minimum density in (\ref{eqn: bounded_density}) and \ref{asmp: cluster_separation} combined are similar to the $(\sigma,\epsilon)$-saliency of \cite{chaudhuri2010}, a standard density clustering assumption, while \ref{asmp: low_noise_density} is seen in, for instance, \cite{singh2009}, (as well as many other works on density clustering and level set estimation.) It is worth highlighting that these assumptions are all local in nature, a benefit of studying a local algorithm such as PPR.

Now we make the crucial distinction between the assumptions made in density clustering, and the stronger conditions we will require here. First, we introduce
\begin{align*}
\mathbf{\Phi}(\sigma, \lambda,\lambda_{\sigma},\gamma) & := \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - \frac{\sigma^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} \\
\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D) & := \frac{\sigma^d}{D^d} \frac{\lambda_{\sigma}^7}{\Lambda_{\sigma}^7} \frac{1}{\log(\Lambda_{\sigma}^2/(\lambda_{\sigma} \sigma^d \nu_d))}.
\end{align*}

Well-conditioned density clusters satisfy all of the given assumptions, for parameters which results in `good' values of $\Phibf$ and $\Psibf$.
\begin{definition}[Well-conditioned density clusters]
	For $\lambda > 0$ and $\Cset \in \Cbb_f(\lambda)$, let $\Cset$ satisfy \ref{asmp: bounded_density} - \ref{asmp: cluster_diameter} with respect to parameters $\sigma, \lambda_{\sigma}, \gamma > 0$ and $\Lambda_{\sigma}, D < \infty.$ Letting $\kappa_1(\Cset)$ and $\kappa_2(\Cset)$ be given by
	\begin{align*}
	\kappa_1(\Cset) & := \frac{\mathbf{\Phi}(\sigma, \lambda,\lambda_{\sigma},\gamma)}{\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D)} \\
	\kappa_2(\Cset) & := \kappa_1(\Cset) \cdot \sqrt{\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D)},
	\end{align*}
	we call $\Cset$ a \textrm{$(\kappa_1, \kappa_2)$-well-conditioned density cluster (with respect to $\sigma, \lambda_{\sigma}, \gamma, \Lambda_{\sigma}$ and $D$).}
\end{definition}

As the notation suggests, $\Phibf$ and $\Psibf$ are related to the graph connectivity criteria. In fact, as we will see in the next section, for well-conditioned density clusters they are exactly (up to constants) the bounds we derive.

@@

Algorithm \ref{alg: ppr} will be the simple variant of PPR we analyze. It will take as input the data $\Xbf$ along with user-specified parameters $r, \alpha$, $\vol_0$, and $v \in \Xbf$.
